---
title: "RNA-seq Analysis"
author: "Ana Romani"
date: "2023-12-01"
---
Information of input: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4878611/

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Call libraries
```{r,echo=FALSE,warning=FALSE}
library("DESeq2")
library("GenomicFeatures")
library("ggplot2")
library("dplyr")
```

## Input upload and check the information
```{r DESeqDataSet, warning=FALSE}
readcounts = read.csv("readcounts_file.csv")
rownames(readcounts)= readcounts[,1]
readcounts = readcounts[,-1]
sample_info <- DataFrame(condition = gsub("_[0-9]+", "", names(readcounts)),
                          row.names = names(readcounts) )
DESeq.ds <- DESeqDataSetFromMatrix(countData = readcounts,
                              colData = sample_info,
                              design = ~ condition)
DESeq.ds
head(counts(DESeq.ds))
```
## Exploratory analysis of the raw counts
We are gonna sum the reads of each gene of each samples to know about the depth of them
How many reads were counted for each sample ( = library sizes)?

```{r eval=TRUE, echo=TRUE}
colSums(counts(DESeq.ds))
```

```{r eval=TRUE, echo=TRUE}
colSums(counts(DESeq.ds)) %>% barplot
```

Remove genes with no reads.

```{r eval = TRUE}
keep_genes <- rowSums(counts(DESeq.ds)) > 0
dim(DESeq.ds)
```

```{r}
DESeq.ds <- DESeq.ds[ keep_genes, ]
dim(DESeq.ds)
```

Now that we have the data in the right format, we can use `estimateSizeFactors()` as the first step to normalize
the data taking in count the difference depth of the samples
Deseq2 uses the method median-of-ratios to calculate the size factors

```{r sizeFactors}
DESeq.ds <- estimateSizeFactors(DESeq.ds)
sizeFactors(DESeq.ds)
```

Check section 5.1.2 of the course notes to see the code for calculating the size
factors yourself with base R functions.

```{r sizeFactor_vs_librarySizes, eval=TRUE, echo=TRUE}
plot(sizeFactors(DESeq.ds), colSums(counts(DESeq.ds)))
```

The read counts normalized for sequencing depth can be accessed via
`counts(DESeq.ds, normalized = TRUE)`.

Let's check whether the normalization helped to adjust global differences between the samples.

```{r boxplots_untransformed, fig.width = 10, fig.height = 5}
# setting up the plotting layout
par(mfrow=c(1,2))
counts.sf_normalized <- counts(DESeq.ds, normalized=TRUE)

# adding the boxplots
boxplot(counts.sf_normalized, main = "SF normalized")
boxplot(counts(DESeq.ds), main = "read counts only")
```
## Exploring the data with normalize counts and compared with raw counts.
```{r boxplots_logReadCounts, fig.width = 10, fig.height = 5, warning=FALSE}
par(mfrow=c(1,2)) # to plot the two box plots next to each other
boxplot(log2(counts(DESeq.ds)), notch=TRUE,
        main = "Non-normalized read counts\n(log-transformed)",
        ylab="read counts")
boxplot(log2(counts(DESeq.ds, normalize= TRUE)), notch=TRUE,
        main = "Size-factor-normalized read counts\n(log-transformed)",
        ylab="read counts") 
```

```{r logReadCountTables}
# non-normalized read counts plus pseudocount
assay(DESeq.ds, "log.counts") <- log2(counts(DESeq.ds, normalized = FALSE)+1)
# normalized and log2-transformed read counts
assay(DESeq.ds, "log.norm.counts") <- log2(counts(DESeq.ds, normalized=TRUE)+1)
```

```{r scatterplots_logNormReadCounts, fig.width = 10, fig.height = 15}
par(mfrow=c(2,1)) 
DESeq.ds[, c("WT_1","WT_5")] %>% assay(., "log.counts") %>% 
    plot(., cex=.1, main = "WT_1 vs. WT_5 (Raw)")
DESeq.ds[, c("WT_1","WT_5")] %>% assay(.,  "log.norm.counts") %>% 
    plot(., cex=.1, main = "WT_1 vs. WT_5")
```
```{r rlog_vs_log2, fig.width = 15, fig.height = 8}
par(mfrow=c(1,2)) 
plot(assay(DESeq.ds[,1:2],  "log.norm.counts"), cex=.1,
     main = "size factor and log2-transformed")
DESeq.rlog <- rlog(DESeq.ds, blind = TRUE)
# the rlog-transformed counts are stored in the accessor "assay"
plot(assay(DESeq.rlog)[,1],
     assay(DESeq.rlog)[,2],
     cex=.1, main = "rlog transformed",
     xlab = colnames(assay(DESeq.rlog[,1])),
     ylab = colnames(assay(DESeq.rlog[,2])) )
rlog.norm.counts <- assay(DESeq.rlog)
```

## Correlation between the counts
```{r }
rlog.norm.counts <- assay(DESeq.rlog)
corr_coeff <- cor(rlog.norm.counts, method = "pearson")
as.dist(1-corr_coeff, upper = TRUE) %>% as.matrix %>%
    pheatmap::pheatmap(., main = "Pearson correlation")
```

## PCA
```{r pca, eval=FALSE}
rv <- rowVars(assay(DESeq.rlog)) # equivalent to rowVars(rlog.norm.counts)
top_variable <- order(rv, decreasing = TRUE)[seq_len(500)]
pca <- prcomp(t(assay(DESeq.rlog)[top_variable, ]))
head(pca$x)
pca_table = as.data.frame(pca$x)
pca_table$condition = as.factor(sample_info$condition)
ggplot(data=pca_table,aes(x=PC1,y=PC2,color=condition))+
  geom_point(size=4,alpha=0.6)+
  theme_classic()
plot(pca$x,col=as.factor(sample_info$condition))
```

## Differential expression analysis
```{r pca, eval=FALSE}
DESeq.ds = DESeq(DESeq.ds)
res <- results(DESeq.ds, contrast = c("condition", "SNF2", "WT"))
res
```
```{r}
resultsNames(DESeq.ds) # tells you which types of values can be extracted with results()
DGE.results <- results(DESeq.ds,
                       independentFiltering = TRUE,
                       alpha = 0.05)
head(DGE.results) # first line indicates which comparison was done for the log2FC
summary(DGE.results)

# the DESeqResult object can basically be handled like a data.frame
table(DGE.results$padj < 0.05)
```

```{r MAplots}
plotMA(DGE.results, alpha = 0.05, 
       main = "Test: p.adj.value < 0.05", ylim = c(-4,4))
```
```{r adjpvalueHistogram}
hist(DGE.results$padj, 
  col="grey", border="white", xlab="", ylab="",
  main="frequencies of adj. p-values\n(all genes)")
```
A sorted results table so that we can immediately see which genes come up as the best candidates:

```{r Deseq}
DGE.results.sorted <- DGE.results[order(-DGE.results$log2FoldChange),]
head(DGE.results.sorted)
```

Plotting counts for single genes (seq. depth normalized, log2-transformed)

```{r singleGenePlot}
par(mfrow=c(1,2))
plotCounts(DESeq.ds, gene="YAL056W", normalized = TRUE)  
plotCounts(DESeq.ds, gene=which.max(DGE.results$padj), main = "Max. p.adj.")
```

```{r heatmap genes comparison}
# identify genes with the desired adjusted p-value cut-off
DGEgenes <- rownames(subset(DGE.results.sorted, padj < 0.05))[1:10]

# extract rlog-transformed values of DE genes into a matrix
rlog.dge <- DESeq.rlog[DGEgenes,] %>% assay

library(pheatmap)
# heatmap of DEG sorted by p.adjust
pheatmap(rlog.dge, scale="none", show_rownames = TRUE,
  main = "DGE (no scaling)")
pheatmap(rlog.dge, scale="row", show_rownames = TRUE,
  main = "DGE (row-based z-score)")
```
```{r singleGenePlot_noDGE, echo = TRUE}
plotCounts(dds = DESeq.ds,
           gene = "YOR290C",
           normalized = TRUE, transform = FALSE,
           main = expression(atop("Expression of "*italic("snf2"), "(YOR290C)")))

```
## GSEA

```{r GSEA analysis}
library(clusterProfiler)
library("org.Sc.sgd.db")
resGSEA <- DGE.results[order(-DGE.results$log2FoldChange),] 
gene_list <- resGSEA$log2FoldChange
original_gene_list <- resGSEA$log2FoldChange
names(gene_list) <- rownames(resGSEA) #Le da el mismo nombre de columnas que nuestros genes
gene_list
gse <- gseGO(gene_list, ont = "ALL", keyType = "ORF", OrgDb = "org.Sc.sgd.db", eps = 1e-300)
dotplot(gse, showCategory=10, split=".sign", font.size=5) + facet_grid(.~.sign)
cnetplot(gse, categorySize="pvalue", foldChange=gene_list, showCategory = 6)
```
```{r}
## clusterProfiler requires a sorted vector where the values correspond
## to the measure used for the sorting 
DGE.results <- DGE.results[order(-1*DGE.results$log2FoldChange),]
genes_for_cp <- DGE.results$log2FoldChange
names(genes_for_cp) <- row.names(DGE.results)

## run the gene set enrichment analysis
gsea_kegg <- clusterProfiler::gseKEGG(geneList = genes_for_cp, organism = 'sce',
                                      nPerm = 1000, minGSSize = 10,
                                      pvalueCutoff = 1, verbose = FALSE)
```

**Dot plots**:

Dot plots depict the enrichment scores and gene counts per gene set (for the most significant gene sets).

```{r dotplot_kegg}
dotplot(gsea_kegg,font.size=6)
```

```{r cnetplot_kegg, message = FALSE, warning = FALSE, fig.width = 16, fig.height = 14}
 cnetplot(gsea_kegg,
   showCategory = 2, node_label = "gene",colorEdge = TRUE, foldChange = genes_for_cp) +
  scale_colour_gradient2(name = "log2FC",
    low = "navyblue", high = "red", mid = "white")
```
# References

Piper, M. M. R. K. M. (2017, April 26). Count normalization with DESeq2. Introduction to DGE. https://hbctraining.github.io/DGE_workshop/lessons/02_DGE_count_normalization.html
GitHub - friedue/course_RNA-seq2019. GitHub. https://github.com/friedue/course_RNA-seq2019/tree/master
GitHub - EveliaCoss/RNASeq_Workshop_Nov2023: Workshop: Análisis de datos de RNA-Seq del 6 al 10 de nov 2023. GitHub. https://github.com/EveliaCoss/RNASeq_Workshop_Nov2023/tree/main

